%!TEX root = ../template.tex
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% acknowledgements.tex
%% NOVA thesis document file
%%
%% Text with acknowledgements
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\typeout{NT FILE acknowledgements.tex}%

\begin{ntacknowledgements}
I would like to begin by expressing my deepest gratitude to my thesis advisors.
Professor Carmen Morgado was the very embodiment of stability, providing invaluable guidance and unwavering support throughout the entire process. Her encouragement, resources, and connections ensured that this thesis could grow with steadiness and clarity. She also directed me to Professor André Lamurías, whose expertise in the field proved essential. His architectural suggestions, critical insights, and thoughtful discussions were pivotal in shaping the scientific rigour of this work.

I am also grateful to NOVA University Lisbon for welcoming me into its academic community, and for the enriching experience I had across all my courses. I would like to thank the professors I had the privilege of learning from, whose engaging lectures and dedication to teaching left a lasting impact on my academic journey.

Beyond the academic sphere, I would like to express my heartfelt thanks to those closest to me.
To my father, whose leading role in our family has inspired me to strive to become a better man and a better person, and to value education as the foundation of the future. I thank him for the courage and resilience he has instilled in me. His example of strength, discipline, and integrity has been a guiding force throughout my life, teaching me to face challenges with determination and to embrace responsibility with dignity.
To my mother, for her warmth, her unwavering support, her constant concern, and the love she has given me at every single step of my journey. She has been the one to worry for me in silence, to celebrate each small victory with joy, and to remind me, through her love, that no obstacle was insurmountable.
To them both, I owe this opportunity.

To my girlfriend, the most important pillar of my life, whose presence has been a constant source of strength and hope. She never ceased to believe in me, never doubted my path, and stood by me through every challenge and uncertainty. Each day, her support and love gave me the stability I needed to face the obstacles of this journey. Without her, this academic achievement would not have been possible. She was not only my guiding light in moments of doubt, but also my greatest motivation to persevere and move forward. This thesis, like this journey, is as much hers as it is mine.

Finally, to my academic friends, with whom I shared joy, challenges, and companionship throughout this journey. The collaboration, mutual encouragement, and countless memories we built together made this path not only easier but also more meaningful. For that, I am truly grateful.
\end{ntacknowledgements}